A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-06 tex/sec4-pre-synthesis.tex
[ tex ] Call Soundness and Completeness a theorem

Erasure of a bidirectional binding signature removes mode information and keeps everything else intact; this can be straightforwardly extended by induction to remove mode information from a bidirectional typing derivation and arrive at an ordinary typing derivation, which is soundness.
We can also remove typing and retain mode information, arriving at a mode derivation instead.
Conversely, if we have both mode and typing derivations for the same term, we can combine them and obtain a bidirectional typing derivation, which is completeness.
In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, which is directed by the same raw term.
%Formulating the notion of mode derivation helps to complete this clean picture.
Technically, all these can be summarised in one lemma.

\begin{lemma}
There is a bidirectional typing derivation of\/ $\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if there are mode and typing derivations of\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$.
\end{lemma}

>>>>

Erasure of a bidirectional binding signature removes mode information and keeps everything else intact; this can be straightforwardly extended by induction to remove mode information from a bidirectional typing derivation and arrive at an ordinary typing derivation, which is soundness.
We can also remove typing and retain mode information, arriving at a mode derivation instead.
Conversely, if we have both mode and typing derivations for the same term, we can combine them and obtain a bidirectional typing derivation, which is completeness.
In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, which is directed by the same raw term.
%Formulating the notion of mode derivation helps to complete this clean picture.
Technically, all these can be summarised in one theorem.

\begin{theorem}\label{lem:soundness-completeness}
There is a bidirectional typing derivation of\/ $\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if there are mode and typing derivations of\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$.
\end{theorem}
